---
title: "lab final BIOL 309"
author: "-your name-"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# INSTRUCTIONS

Each group will be assigned a different study from the primary literature. As a group, you will choose one question from the paper to answer with your analysis. Once I have approved your group's question, you may start on the activity.

Use previous class materials to construct and run a PVA, and then examine the output. Your document should contain the following:

1. Answer to your specific question
2. Extinction risk and probability of decline of the population in 50 years

You will work as a group, but each group member will make their own individual document using this script. Click "Knit" to save to PDF or Word document. 

Then answer the questions in the Word document.


-------------------------------------------------------------------------------

# BACKGROUNG INFO

*Fill in the following*

Species:

Location:

Main threats:

Question to answer:


-------------------------------------------------------------------------------

# PARAMETERS

What parameters do you need for your model? 

- Parameters for population dynamics
- Parameters specific to your question

```{r}

```

-------------------------------------------------------------------------------

# SPECIFY THE MODEL

Choose one of the models we used in class to answer your question. Paste it here.

```{r}

```

-------------------------------------------------------------------------------

# RUN THE MODEL

Run the model function, using your parameters as the arguments (the things in parentheses)

```{r}

```

-------------------------------------------------------------------------------

# OUTPUT 

First, choose the output that makes sense to answer your chosen question. What measure of population viability would you use? Paste the appropriate code below to create graphs of this measure.

```{r}

```

```{r}

```


```{r}

```


-------------------------------------------------------------------------------

# EXTINCTION RISK

```{r}

```


-------------------------------------------------------------------------------

# PROBABILITY OF DECLINE

```{r}

```


-------------------------------------------------------------------------------


-------------------------------------------------------------------------------
