<<<<<<< HEAD
---
title: "Population Viability Analysis"
author: "Sarah Bolinger"
date: "2023-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggpubr)
```

# REVIEW

## Before we start, let's make sure everyone is properly connected to GitHub.

----------------------------------------------------------------------------------------
*R CODE TROUBLESHOOTING NOTES*

1. If you get an error about something not existing, check spelling, capitalization, underscores, etc. Remember that names need to be typed *identically* every time so R knows which object you're referring to! 

2. Make sure you've loaded the necessary *packages* and imported all necessary *data*. If a necessary package isn't installed, RStudio should prompt you to install it, but you can also install manually. 

3. R doesn't know when you've typed a name incorrectly or failed to load a package, but it will generally tell you when your coding syntax (formatting, essentially) is off with a little red x in front of the line where the problem is. If you mouse over the x symbol, a message pops up telling you what the issue is (some of these are easier to understand than others). Things to watch for: close your parentheses and quotes, don't add extraneous symbols or letters, add commas between function arguments, put all non-numeric writing in quotations.

4. Throughout the script, I will put extra notes about R in brackets [] for anyone who is interested in knowing why things look the way the do. Feel free to ignore them for now if you're feeling overwhelmed!

5. If you are antsy waiting for one of us to come help you, try using Google to find an answer! This is also good practice for when you are working on your final projects in R later this semester. You can try copying the exact text of an error message if you don't know what to search. Also try searching for "R + ~name of function~" for more info about things.
----------------------------------------------------------------------------------------

## Now, some review of what we did in our last few R labs:

### In R Markdown, you have *text* sections (like this one) and *code chunks* (which should have a different background color). The text sections are almost like a Word document, so you can type regular text. The formatting uses special symbols, but we don't need that for now.

[R NOTE: You can make a new empty code chunk by typing ctrl+alt+i ]

```{r}
# code chunk
# non-code needs to be commented, like this (using pound sign)
if it isn't, R will try to read it as code and will get mad that the syntax isn't correct
```

### Normally at the beginning of a script we load in packages. We loaded "tidyverse" and "ggpubr" above.

[R NOTE: what are the ## in the *text* (not the code chunks) telling R to do?]

### R has built-in data we can use. Let's load the data so we can view it. Find the dataframe under "Environment". Now click on the arrow next to it to see the variables it contains. Alternatively, click on the name itself to open a view of the whole dataframe (then close it and come back to the script).

```{r}
data("fish_encounters")
```

### This data represents captures of fish at different stations as they swim downstream. The stations are in order from upstream (Release) to furthest downstream (MAW). Let make a table of the number of captures at each station. What happens as we get to stations further downstream?

```{r}
table(fish_encounters$station)
```

### Now let's make a frequency distribution of the same data - the number of fish caught at each station. Note that the two ways of writing the plot code below produce identical results. If you prefer putting name of the data within the function call, you can do that instead of "piping" it in. Piping helps make the code more readable because it puts the name of the data first, so you may prefer that instead.

```{r}
ggplot(data=fish_encounters, aes(x=station)) +
  geom_bar()

fish_encounters |> ggplot(aes(x = station)) +
  geom_bar()
```

A simple way to visualize captures of each individual as they move downstream is to make a bar chart with a different color for each individual fish:

```{r}
fish_encounters |> ggplot(aes(x = station)) +
  geom_bar(aes(fill=fish)) +
  # all of the code below is just to get the legend looking nice!
  theme(legend.text = element_text(size=7),
        legend.position = "top", 
        legend.key.width = unit(0.3, units="cm")) +
  guides(fill = guide_legend(nrow=1, label.position="bottom"))
```


## Let's import some external data. 

### Note the name of the dataframe (capt) and view the variables like we did above.

```{r}
# ufo <- read.csv("ufo_location_shape.csv")
# ufo<- read.table(file=url("https://users.stat.ufl.edu/~winner/data/ufo_location_shape.csv") )
ufo<- read_csv(file=url("https://users.stat.ufl.edu/~winner/data/ufo_location_shape.csv") )
```

How would you make a simple table to show the total number of UFO sightings per country?

```{r}

```


Change the plot below to show the frequency of different *shapes* of UFO that have been reported:

```{r}
ggplot(ufo, mapping = aes(x=State)) +
  geom_bar() +
  coord_flip() +
  theme(axis.text.x=element_text(size=5))
```

# DATA FROM LAST LAB

You can create a csv (essentially a small text-based version of a spreadsheet) directly in R, so you don't even have to open Excel or Google Sheets.

We are going to enter our data by making each *column* on the datasheet into a *vector* in R (vector = kind of like a list, but easier to manipulate in R). 

Then, we will join all the vectors together to create our dataframe. We could also do this by row ("row-wise") and not column-wise, but I think it's easier to visualize the columns because each one represents a variable.

Enter each value from each column, separated by commas (we'll do an example first)

```{r}
########### EXAMPLE DATAFRAME ####################

Species <- c("CACH", "BRTH", "GRCA", "BRTH")
ID      <- c(1, 2, 3, 4)
Age     <- c("HY", "U", "AHY", "HY")
Sex     <- c("M", "U", "U", "U")
Comments <- c("molting primaries 1 and 2", 
              "none", 
              "the best bird", 
              "eye color is gray- indicates hatch year")

birdData <- data.frame(Species, ID, Age, Sex, Comments)
```

[R NOTE: note how I can add spaces and line breaks within the parentheses and it does not affect how R interprets the code. This can improve readability.]

```{r}
Species <- c()
ID      <- c()
Age     <- c()
Sex     <- c()
Comments <- c()

labData <- data.frame(Species, ID, Age, Sex, Comments)

```


# POPULATION VIABILITY ANALYSIS SIMULATION

*Population Viability Analysis (PVA)* is one of the most commonly used management practices. It takes parameter estimates (i.e., estimated numbers for processes that we care about, like birth rates, death rates, immigration & emigration) from the published literature to make predictions about what will happen to the population under different scenarios. 

We are implementing a very simple PVA here, but more complex analyses can account for the degree of genetic variability, susceptibility to diseases, and other relevant measures of population health, or viability. Today, you are going to practice going through the analysis and learning what the parts do. 

*NOTE* We will be using this model again in the future! For your final project, you will search the published literature for estimates of the parameters given here (e.g., R_max, Init_N), and put those into the model to perform a PVA for your chosen species.

*You will want to run through the whole script once before answering questions on the lab report.* 

There are packages for running PVA analyses, but you will often find it easier to troubleshoot if you have the code in front of you (even if you found it online and didn't write it yourself). Instead of just telling R "run this analysis," you can see what goes into the analysis and make sure it's doing what you want (and change it as needed).


## MODEL PARAMETERS 

Before running the model, we will tell R what values to use for different parameters.

**Parameters** is just a fancy way of saying "values we might change within the model" Some examples of parameters used in population models are:  


*population rates* (e.g. growth rate, mortality rate, immigration rate)  

*initial values* (here, the initial population size)

*stochasticity* (values we choose to represent randomness in the environment or between individuals)

*external factors* (e.g. probability of flood, storm, hunting, etc.)

*population processes* (e.g. carrying capacity, density dependence)


But remember that not all models will use all of these. Example: migration is difficult to measure in most species, and we may not have estimates. Or we may choose to create a simplified model depending on the question we are asking.


### Basic life history parameters

```{r}
R_max <- 1.05       # Maximum rate of growth
Init_N <- 50        # Initial abundance
K <- 175            # Carrying capacity
```

### Environmental stochasticity (randomness)

The lambda parameter is the population growth rate in one time step (Nt+1/Nt).

```{r}
SD_lambda <- 0.11  # standard deviation of lambda
```


### Density-dependence (Ricker model)

This is a function for computing next-year abundance -- includes environmental stochasticity. It is conceptually similar to the logistic growth rate we have seen in lecture.

```{r}

NextYear <- function(prev_abund){       
  prev_abund * exp(log(rnorm(1,R_max,SD_lambda))*(1-(prev_abund/K)))
}

```

### Catastrophe

Every year, there is some chance of flood occurrence, and the severity of the flood influences how many individuals will survive

```{r}
Flood_prob <- 0.05     # 5% chance of major flood
Flood_surv <- 0.9      # 90% of population can survive a flood 
```

### Basic simulation parameters

Tell R how long to run the model for, and how many times to run it.

```{r}
nyears <-100     # number of years
nreps <- 500      # number of replicates
```


## SPECIFY THE PVA MODEL

[R NOTE:If you want to look at the function step-by-step, enable the *browser* function below]

```{r}


PVAdemo <- function(nreps, nyears, Init_N, R_max, K, Flood_prob, Flood_surv){
  # browser()
  PopArray2 <- array(0,dim=c((nyears+1),nreps))
  
  for(rep in 1:nreps){                            # start looping through replicates
    
    PopArray2[1,rep] <- Init_N                               # initial abundance
    

    for(y in 2:(nyears+1)){                                  # loop through years

      nextyear <- max(0,trunc(NextYear(PopArray2[y-1,rep]))) # stochasticity + density-dep
      
      if(runif(1)<Flood_prob) nextyear <- nextyear*Flood_surv       # catastrophe
        
      PopArray2[y,rep] <- nextyear 
      
    }
  }
  
  return(PopArray2)
}

# PVAdemo()
```


## RUN THE PVA MODEL

```{r, warning=FALSE}

Default <- PVAdemo(nreps,nyears,Init_N,R_max,K,Flood_prob, Flood_surv)

# make output easier to view and add year column
Default <- data.frame(year = seq(1, nyears + 1, by=1), Default) 
Default[is.na(Default)] <- 0


```


## PLOT THE MODEL OUTPUT

There are several types of visualizations that you might want to use for your PVA models: The first is to look at the "cloud" of population abundance trajectories. What is this showing us?

We will first specify a function to plot the data, and then run the function. This means we don't have to type out the plot code multiple times (we can just plug in the name of the data to plot)

```{r}
plot_cloud <- function(simdata){
  
  dat <- vector("list", nreps)
  p <- ggplot()
  time <- seq(0, nyears, by = 1)
  
  for (i in seq(2,nreps)) {
    s <- simdata[,i] 
    dat[[i]] <- data.frame(t = time, s = s)
    p <- p + geom_line(data = dat[[i]], mapping = aes(x = t, y = s), col = i)
  } 
  
  return(p)
}
p <- plot_cloud(Default)
p #or print(p)
```


## OTHER ANALYSES

We can create other functions to answer other questions of interest using our simulated data.

Maybe our question is about the **probability of decline over some time period** 

(Note: This is VERY common in management -- the way managers evaluate endangered species listings is by determining what the population will look like in 30 years, 60 years, etc.).  

In that case maybe we should present a histogram of final abundances (in other words, **looking at all the model runs, how many times did the population size at year 100 end up in each of the following bins?**):

[R NOTE: This time we aren't making a function to make the plot, we're just making the plot]

```{r}

hist(as.numeric(Default[nrow(Default),]),xlab="Final population size at endpoint",ylab="Number of replicates",main="")
# abline(v=Init_N,col="blue",lwd=2)



```


Or if our question is about extinction risk, maybe we want to plot extinction risk by time??? Create the function:

```{r}
Extinction_byyear <- function(simdata){
  apply(simdata,1,function(t)  length(which(t==0)))/ncol(simdata)
}
```

Run the function and plot:

```{r}

ext <- data.frame(year = seq(0, nyears, by=1),ext = Extinction_byyear(Default)) 

ext |> ggplot(aes(x=year, y=ext)) +
  geom_line() +
  ylab("Extinction Risk")

```

Why is the plot of extinction risk like this? Check to see if the population went extinct in any of the model runs. 

What if our question is about the effect of flooding on extinction risk. Let's imagine that the probability of flooding is not expected to change with climate change, but that the intensity of the flood damage is likely to increase substantially.  Currently, floods generally result in a 10% population reduction (90% of the population survives the flood). But climate change could increase this number to as much as 85%. Let's look at how much this could increase extinction risk.

Specify the function:

```{r}

Exctinction_risk <- function(simdata){
  length(which(simdata[nrow(simdata),]==0))/ncol(simdata)
}
```

Since we don't know the amount of population reduction caused by climate change, we estimate. Currently, there is a 10% population decline (lambda (Flood_surv) = 0.9). The literature says climate change could lead to up to 85% declines (Flood_surv =0.15). **The uncertainty in the estimate means we have to allow for a range of future conditions**. 

```{r}
flood_survivors <- seq(0.9,0.15,by=-0.05) # list of possible flood survival probabilities
```

This is exactly the same PVA as before, except now the proportion surviving the flood is not a constant, but a parameter that varies

```{r}

all_scenarios <- numeric(length(flood_survivors))
for(scenario in 1:length(flood_survivors)){
  PVA <- PVAdemo(nreps,nyears,Init_N,R_max,K,Flood_prob, flood_survivors[scenario])
  all_scenarios[scenario] <- Exctinction_risk(PVA)
}
survivors <- data.frame(scenario = all_scenarios, flood_survivors)

ggplot(survivors, aes(x=flood_survivors, y=scenario)) +
  geom_point(shape=17, size=3, col="purple3")  +
  ylab("Extinction Risk") +
  xlab("Proportion of Individuals Surviving Flood") +
  geom_smooth(method="lm",
              formula=y~x) + 
  stat_regline_equation(label.x=0.8,aes(label=..rr.label..)) + 
  stat_regline_equation(label.y=0.8,label.x=0.8,aes(label=..eq.label..)) 
  
# plot(flood_survivors,all_scenarios,type="p",cex=2,xlab="Proportion of Individuals Surviving Floods",ylab="extinction risk")
#abline(h=0.05,col="turquoise",lwd=2)
```


# APPLY TO REAL-WORLD DATA

How would we use the above code on real-world data? Let's try with the data we collected last lab. We have already specified all the functions in R, so we just need to *call the functions* on our lab data.

```{r}

```
